Signatures of Foreground Sky in Power Spectra of Redshifted Neutral Hydrogen from the Epoch of Reionization

### 2014-09-10 foreground_delay_spectrum_modeling.tex
From v0.4 (b4741f4) and 2e40b94 from Danny: ==========================================
(*) Created a branch danny2 and merged Danny's updated repository into this
branch
(*) Removed references to labels sec:discussion and sec:discussion to let the
file compile in LaTeX without warnings and modified those places accordingly.
(*) Minor edits throughout
(*) Changed structure slightly in section 3 to increase reading continuity

Perhaps the key challenge in the statistical detection of power spectrum of the redshifted H{\sc i} signal arises from the contamination by Galactic and extragalactic foregrounds \citep[see, e.g.,][]{dim02,zal04,fur06}. \citet{mor04} show that the inherent isotropy and symmetry of the EoR signal in frequency and spatial wavenumber ($k$) space make it distinguishable from sources of contamination which are isolated to certain $k$ modes by virtue of their inherent spectral smoothness\citep{mor06,bow09,liu11,par12,dil13}. Considerable effort has been made towards understanding the $k$-space behavior foreground signatures in the observed power spectrum \citep{thy13,pob13,mor12,tro12,dat10,bow09}. A general model of how smooth spectrum foregrounds are scattered to higher $k$-modes by the wide-field response of the instrument, into the so-called `wedge', has emerged that provides a high level explanation for the observed foreground power spectrum. The conservative foreground strategy that has developed alongside this work is to discard $k$-modes which could be contaminated (Avoidance). The more aggressive alternative is to subtract a sky model and regain access to modes lost by Avoidance. In both cases, which parts of the sky are most critical to either Avoid or Subtract has remained largely uncertain. Here we will focus primarily on the Avoidance strategy, focusing on identifying foreground components most likely to ``leak'' from foreground modes to EoR modes.

Foregrounds leak into the EoR H{\sc i} power spectrum by either having intrinsic spectral variability, high instrumental chromaticity, or through approximations in power spectrum analysis.  Here we use existing catalogs and a high fidelity instrumental model to focus on the possible sources of instrumental chromaticity. 
To decouple from possible analysis effects, such as those pointed out by \citet{haz13}, we compute power spectra using the individual antenna pair based approach of \citet{pob13,par14}.  This approximates the power spectrum as the inverse fourier transform of the spectra generated by the correlator.

 Matching real observations from the Murchison Widefield Array Epoch of Reionization observing program with instrumental simulations of the entire known sky we examine the signature of the power spectrum of each baseline.  We confirm that the foregrounds most likely to obfuscate the Epoch of Reionization power are not caused by emission central field of view but rather from near the horizon. We also find that the smooth galactic emission plays a larger role then previously thought. We quantify this contamination by breaking the simulations down by sky region and type of emission. Using these we arrive at a new method for minimizing the contribution of bright foregrounds which uses prior knowledge of the sky to downweight badly contaminated baselines.

In \S\ref{sec:delay-spectrum} we introduce the delay spectrum technique, then in  \S\ref{sec:instrument}, \S\ref{sec:obsparms}, and \S\ref{sec:data-analysis} we describe the MWA setup, summarize the observing parameters and present the resulting data.  Simulations of these observing parameters are described in \S\ref{sec:modeling} and analyzed by sky location, and type of emission in \S\ref{sec:delay-spectrum-analysis}.  \S\ref{sec:discussion} offers a discussion along with a basic comparison to other current and future instruments. \S\ref{sec:conclusion} offers conclusions and an initial description of a more precise foreground avoidance technique.

@@

Interferometer array data known as {\it visibilities}, $V_f(\overline{\mathbf{u}},f)$, represent correlations between time--series of electric fields measured by different antenna pairs with separation vectors $\overline{\mathbf{x}}$ and then Fourier transformed along time axis to obtain a spectrum along frequency ($f$) axis. Here, $\overline{\mathbf{u}}\equiv \overline{\mathbf{x}}f/c \equiv \overline{\mathbf{x}}/\lambda$, $c$ is the speed of light and $\lambda$ is the wavelength. $\overline{\mathbf{u}}$ is related to transverse spatial frequency modes ($\overline{\mathbf{k}}_\perp$) of brightness distribution of the sky as $\overline{\mathbf{k}}_\perp \equiv 2\pi\overline{\mathbf{u}}/D(z)$, where $D(z)$ is the transverse comoving distance at a redshift $z$. If $I(\overline{\mathbf{l}},f)$ is the emission at different frequencies on the sky as a function of direction unit vector ($\overline{\mathbf{l}}$) specified as direction cosines, $V_f(\overline{\mathbf{u}},f)$ represents the Fourier decomposition of $I(\overline{\mathbf{l}},f)$ attenuated by the antenna's angular power pattern $A(\overline{\mathbf{l}},f)$, into transverse spatial frequency modes. Taking into account the instrumental bandpass weights $W_f(f)$, this relation can be written as:
\begin{align}\label{eqn:obsvis}
  V_f(\overline{\mathbf{u}},f) &= \int A(\overline{\mathbf{l}},f)\,I(\overline{\mathbf{l}},f)\,W_f(f)\,e^{-i2\pi \overline{\mathbf{u}}\cdot\overline{\mathbf{l}}}\,\dif^2 \overline{\mathbf{l}}.
\end{align}

We define the {\it delay spectrum} $V_\eta(\overline{\mathbf{u}},\eta)$ to be the inverse Fourier transform of $V_f(\overline{\mathbf{u}},f)$ along the frequency coordinate:
\begin{align}\label{eqn:delay-transform}
  V_\eta(\overline{\mathbf{u}},\eta) &= \int V_f(\overline{\mathbf{u}},f)\,W'_f(f)\,e^{i2\pi\eta f}\,\dif f,
\end{align}
where, $W'_f(f)$ is a frequency window weighting function which can be chosen to control the quality of the delay spectrum \citep{thy13,ved12}. $\eta$ represents the signal delay between the antenna pairs in an interferometer given by:
\begin{equation}\label{eqn:delay}
  \eta = \frac{\overline{\mathbf{x}}\cdot\overline{\mathbf{l}}}{c} = \frac{\overline{\mathbf{u}}\cdot\overline{\mathbf{l}}}{f}.
\end{equation}
Note that as the visibility is a fully complex quantity, its Fourier transform is fully Nyquist sampled over positive and negative delays.  Negative and positive delay\footnote{The terms delay and lag are used interchangeably in this paper.} correspond to the two hemispheres of the sky as transected by the antenna pair's baseline vector. For example a perfectly East-West baseline will see a source in the Eastern sky at positive delay, and conversely for a source in the Western sky.

Since we are studying a redshifted H{\sc i} spectral line, $f$ is a measure of cosmological distance along the line of sight. Consequently, $\eta$ is a measure of line of sight spatial frequency mode, $k_\parallel$. Thus, $|V_\eta(\overline{\mathbf{u}},\eta)|^2$ is directly related to the spatial power spectrum of redshifted H{\sc i} distribution, $|V(\overline{\mathbf{k}}_\perp,k_\parallel)|^2$. However, foreground objects emit in radio frequencies and contaminate the signals from the redshifted H{\sc i}. Due to their spatial and spectral properties, $V_\eta(\overline{\mathbf{u}},\eta)$ also contains contamination from foreground radio sources \citep{thy13,tro12,mor12,bow09}. These contaminations have to be characterized precisely in order to reduce their impact on redshifted H{\sc i} power spectrum detection sensitivity.

@@

The MWA was completed in 2012, and after commissioning, began its EoR observing program in 2013. In its final configuration the MWA is a 128--tile interferometer array capable of observing a 30~MHz band anywhere in the range 80--300~MHz.  Each tile is a phased array of 16 dipoles each in the shape of a bow--tie characterized by a grating response with a $\sim$~25\arcdeg wide primary field of view and multiple secondary side lobes. The tile arrangement is a centrally condensed core of $\sim$~300~m --- there are many spacings in the range 5--50~m --- and a radial density that falls off as the inverse of the radius with the longest baselines at 3~km \citep{bea12}. Here we focus on antenna spacings $|\overline{\mathbf{x}}| \le 200$~m (spatial scales relevant to reionization) and a frequency range 170--200~MHz.

@@

The LSTs of these two snapshots are shown in Figure~\ref{fig:pointings} as vertical lines at -1.92~hours (i.e., 22.08~hours) and 0.09~hours which are hereafter denoted as {\it off--zenith} and {\it zenith} pointings respectively. %These pointings (phased array power patterns) are centered at RA = 4\fdg 387, Dec = -29\fdg 86 and RA = -28\fdg 8, Dec = -26\fdg 701, whereas the visibilities themselves are phased to zenith corresponding to RA = -28\fdg 8, Dec = -26\fdg 701, and RA = 1\fdg 35, Dec = -26\fdg 701 respectively.

@@

\subsubsection{Initial Data processing}\label{sec:data-analysis}
The data are flagged\footnote{\url{http://sourceforge.net/p/aoflagger}} for interference \citep{off12,off10}, removing 3\% of the data and averaged in time  and frequency from the raw 1/2s,40kHz to 2s, 80khz. These data are then calibrated to a simulation of the sky containing 2420 point sources selected from the MWA Commissioning Survey (N.H Walker, in prep).  MWACS was compiled using observations taken with subsets of the array during the final buildout and commissioning. It has a flux limit of 25~mJy and a the declination range of -12\arcdeg to -40\arcdeg evenly covering the field of view of the observations reported here.        The calibration sources are selected to fall within the 5\% primary lobe of the tile beam. The calibration algorithm computes a per channel, per antenna, complex gain solutions averaged to two minute intervals.  The solutions are still fairly low signal to noise so we iteratively project along the antenna and frequency dimensions to capture the relatively independent passband and antenna to antenna variation. First we average the channel gains over all antenna to obtain a high signal to noise measurement of the bandpass.  After applying this single passband, we do a second round of calibration and fit a 2nd order amplitude and first order phase for each antenna. This flattens any residual variation in bandpass and removes small phase slopes due to variation in cable delay. Finally we fit for an additional phase known to be caused by reflection in 150m cables. %possibly could add elaboration, if I knew exactly which version of the cal was done here. doesn't really matter

@@

We obtain the delay spectrum of these calibrated visibilities taking the delay transform of each baseline's spectrum (equation~\ref{eqn:delay-transform}) choosing $W'_f(f)$ to be a {\it Blackman--Harris} window function. The spectrum is multiplied by  the additional weight of the flagged channels which occur rarely for interference and at at regular spacing every 1.28~MHz where the edges of coarse channels are known to have a slight aliasing effect.  In delay space, these weights translate into a convolution by a ringing harmonic point spread function (PSF).  We deconvolve this PSF using a one dimensional CLEAN algorithm \citep{tay99} as described for the delay axis by \citet{par09,par12}. The CLEAN procedure iteratively finds and subtracts peak values convolved by the Fourier transform of the weights.   We limit the selection of peaks to modes inside the delay horizon corresponding to smooth spectrum sources on the sky --ie beneath the black line in Figure \ref{fig:fhd_data}.

@@

We show results of delay spectrum from MWA data after deconvolution along delay axis in figure~\ref{fig:fhd_data}. In this plot and others, we plot the amplitude of the delay spectrum sorted by baseline length. Note that no baselines have been averaged together in this plot.  The principle features are the wedge-shaped feature the boundary of which increases approximately linearly with light travel time of the baseline length.  Note that the wedge is not filled out uniformly. Each baseline has a different orientation and resulting response to different sources in the sky. It is this variation that we are investigating.

In these power spectrum units [Mpc$^{-3}$], the reionization power spectrum is expected to be spherically symmetric and decrease quickly with $|k|$.  The brightest signal will occur on the shortest baselines and smallest delays. Thus the region of interest is just above the black horizon line and below the harmonic due to the coarse channels.

 The {\it off--zenith} pointing has emission notably higher than in the {\it zenith} pointing inside the wedge boundaries. This is shown later to be due to response of the array to the bright Galactic center and Galactic plane in the westward sky. Consequently, the contamination into the {\it EoR window} is also higher. Emission in the {\it zenith} pointing is more centrally concentrated due to phasing of power pattern to the zenith, whereas the {\it off--zenith} pointing has a power pattern phased eastward of zenith and is also responsive to emission from angles far away from zenith.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/multi_baseline_fhd_delay_spectrum_snapshots.eps}
\caption{ Delay spectrum power spectra of two 112s  Murchison Widefield Array observations of the Epoch of Reionization field at RA=0, Dec=-30.  At top an {\it off--zenith}  pointing when tiles have significant side lobe response in the direction of the galactic center. At bottom a {\it zenith} pointing where side lobes are minimal and the galaxy has completely set.  To get the the power spectrum each baseline is Fourier transformed into delay space ($y$-axis) and the amplitude plotted as a logarithmic color in order of baseline length ($x$-axis). The logarithmic color scale (shown at the right) is common to both panels. White lines mark the boundaries of the {\it foreground wedge} determined by the light travel time across each baseline. Smooth spectrum sources are physically limited to the region below this line.  The various features within the wedge depend on the \emph{orientation} of the baseline, the side lobe structure of the primary beam pointing and the sky at that time. \label{fig:fhd_data}}
\end{figure}

Despite delay deconvolution, we still see leakage beyond the cleaned regions indicating the inability of the deconvolution algorithm to fit the data perfectly. This is predominant at short antenna spacings where the number of delay modes within the horizon. This is because the maximum delay envelope (boundary of the wedge) consists of mixed emission from large portions of the sky into a narrow range of delays and the deconvolution algorithm does not have sufficient support in delay space to fit the delay spectrum. However, we wish to emphasize that the purpose of deconvolution is to rid the delay spectra of instrumental artifacts to the best extent possible in order to see the underlying signatures of foreground emission. It is not our intention in this paper to use the deconvolution as a foreground removal tool. Hence, our discussion and conclusions are valid despite imperfections in deconvolution.

@@

We describe power pattern and foreground models we have used in modeling the instrument. %todo

\subsection{Power pattern}\label{sec:power_pattern}

The precise power pattern of the MWA tile ($A(\overline{\mathbf{l}},f)$ in equation~\ref{eqn:obsvis}) is a subject of active investigation. The most precise model currently available is that of Sutinjo et al (2014; submitted?) which models the 4$\times$4 array of bow-tie dipoles as a mutually-coupled array with the overall power-pattern of the dipole modeled via finite element electromagnetic simulation.  To speed simulation we find that a phased array of isotropic radiators at a height of 0.3~m above an infinite ground plane provides a very good approximation to the FEKO simulation. We also assume that each individual dipole signal has random delay fluctuations of rms 0.05~ns, a number in line with the known stability level of the analog signal chain.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/delta_array_powerpattern_0.3m_ground_custom.eps}
\caption{Models of MWA tile power pattern in Right Ascension ($\alpha$) and Declination ($\delta$) coordinates for the off--zenith (top) and zenith pointings (bottom) at 185~MHz. An MWA tile is modeled as a 4$\times$4 array of isotropic radiators placed 0.3~m above the ground plane. Random fluctutations of rms 0.05~ns have been added to delay corrections during the phased addition of voltages from these isotropic radiators. The logarithmic color scale (shown at the right) is common to both panels. \label{fig:power_pattern}}
\end{figure}

\subsection{Foreground Model}\label{sec:foreground}
%todo
The MWA has a very wide field of view ($\gtrsim 20$\arcdeg at 185~MHz) and has non-zero response all the way to the horizon. As we have seen above, foreground sources contaminate the region inside the ``wedge'' with an amplitude that depends on the primary beam amplitude in that direction, with sources at the horizon being closest to the edge of the wedge.

The degree to which these sources leak power into spectral line of sight modes above the wedge, a region known as the ``EoR Window'', depends primarily on the brightness of foregrounds near the edge of the wedge \citep{thy13,pob13,ved12,par12}. Thus, it is important to consider an all-sky model for foreground objects in evaluating the features seen in the power spectrum instead of restricting to the primary field of view.

\citet{bea13} estimated the sensitivity of the MWA to EoR H{\sc i} power spectrum detection taking into account thermal noise effects in the {\it Eor window}. \citet{thy13} estimated the sensitivity taking into account the spillover from foreground contamination from unsubtracted extragalactic point sources in the {\t EoR window} besides thermal noise. In this paper, we include the diffuse Galactic emission for enhancing our understanding of foreground signatures in the power spectrum. We use an all-sky foreground emission model that includes both diffuse and bright compact components.

@@

For the diffuse component, we use an all-sky radio foreground model \citep{deo08} to estimate the emission at 185~MHz. Since this map is predominantly based on the 408~MHz all-sky map of \citet{has82} which has an angular resolution of 0\fdg 85, we smoothed the 185~MHz diffuse emission map to the same resolution. However, to avoid any artifacts from sampling this map, we sample it at $\approx 27$\arcmin~intervals. Using the same model, we use the diffuse emission maps at 170~MHz and 200~MHz to estimate the spectral index map of the diffuse emission model.

A low resolution version of the diffuse foreground model used is show in figure~\ref{fig:DSM}. Contours of the MWA tile power pattern shown in figure~\ref{fig:power_pattern} are also overlaid. Of notable significance in the {\it off--zenith} pointing is the presence of a portion of the Galactic plane and the bright Galactic center in the visible hemisphere in the westward sky and the MWA tile's power receptivity is significantly high ($\gtrsim 12$\%) in that direction. In the {\it zenith} pointing, the Galactic plane has almost set and the contour level of power pattern in that direction is at least 16 times smaller.

>>>>

Perhaps the key challenge in the statistical detection of power spectrum of the redshifted H{\sc i} signal arises from the contamination by Galactic and extragalactic foregrounds \citep[see, e.g.,][]{dim02,zal04,fur06}. \citet{mor04} show that the inherent isotropy and symmetry of the EoR signal in frequency and spatial wavenumber ($k$) space make it distinguishable from sources of contamination which are isolated to certain $k$ modes by virtue of their inherent spectral smoothness \citep{mor06,bow09,liu11,par12,dil13}. Considerable effort has been made towards understanding the $k$-space behavior foreground signatures in the observed power spectrum \citep{thy13,pob13,mor12,tro12,dat10,bow09}. A general model of how smooth spectrum foregrounds are scattered to higher $k$--modes by the wide-field response of the instrument, into the so-called `wedge', has emerged that provides a high level explanation for the observed foreground power spectrum. The conservative foreground strategy that has developed alongside this work is to discard $k$--modes which could be contaminated ({\it avoidance}). The more aggressive alternative is to subtract a sky model and regain access to modes lost by {\it avoidance}. In both cases, which parts of the sky are most critical to either {\it avoid} or {\it subtract} has remained largely uncertain. Here we will focus primarily on the {\it avoidance} strategy, focusing on identifying foreground components most likely to ``leak'' from foreground modes to EoR modes.

Foregrounds leak into and contaminate the EoR H{\sc i} power spectrum by either having intrinsic spectral variability, high instrumental chromaticity, or through approximations in power spectrum analysis. Here we use existing catalogs and a high fidelity instrumental model to focus on the possible sources of instrumental chromaticity. 
To decouple from possible analysis effects, such as those pointed out by \citet{haz13}, we compute power spectra using the individual antenna pair based approach of \citet{pob13,par14}. This approximates the power spectrum as the inverse fourier transform of the spectra generated by the correlator.

 Matching real observations from the Murchison Widefield Array Epoch of Reionization observing program with instrumental simulations of the entire known sky we examine the signature of the power spectrum of each baseline. We confirm that the foregrounds most likely to obfuscate the Epoch of Reionization power are not caused by emission central field of view but rather from near the horizon. We also find that the smooth galactic emission plays a larger role then previously thought. We quantify this contamination by breaking the simulations down by sky region and type of emission. Using these we arrive at a new method for minimizing the contribution of bright foregrounds which uses prior knowledge of the sky to down--weight adversely contaminated baselines.

In \S\ref{sec:delay-spectrum} we introduce the delay spectrum technique, then in  \S\ref{sec:instrument}, we describe the MWA setup, summarize the observing parameters and present the resulting data. Simulations of these observing parameters are described in \S\ref{sec:modeling} and analyzed by sky location, and type of emission in \S\ref{sec:delay-spectrum-analysis}. In \S\ref{sec:fg-grading}, we offer an initial description of a more precise foreground avoidance technique. \S\ref{sec:summary} summarizes our work and findings.

@@

Interferometer array data known as {\it visibilities}, $V_f(\overline{\mathbf{u}},f)$, represent correlations between time--series of electric fields measured by different antenna pairs with separation vectors $\overline{\mathbf{x}}$ and then Fourier transformed along time axis to obtain a spectrum along frequency ($f$) axis. Here, $\overline{\mathbf{u}}\equiv \overline{\mathbf{x}}f/c \equiv \overline{\mathbf{x}}/\lambda$, $c$ is the speed of light and $\lambda$ is the wavelength. $\overline{\mathbf{u}}$ is related to transverse spatial frequency modes ($\overline{\mathbf{k}}_\perp$) of brightness distribution of the sky as $\overline{\mathbf{k}}_\perp \equiv 2\pi\overline{\mathbf{u}}/D(z)$, where $D(z)$ is the transverse comoving distance at a redshift $z$. If $I(\overline{\mathbf{l}},f)$ is the emission at different frequencies on the sky as a function of direction unit vector ($\overline{\mathbf{l}}$) specified as direction cosines, $V_f(\overline{\mathbf{u}},f)$ represents the Fourier decomposition of $I(\overline{\mathbf{l}},f)$ attenuated by the antenna's angular power--pattern $A(\overline{\mathbf{l}},f)$, into transverse spatial frequency modes. Taking into account the instrumental bandpass weights $W_f(f)$, this relation can be written as:
\begin{align}\label{eqn:obsvis}
  V_f(\overline{\mathbf{u}},f) &= \int A(\overline{\mathbf{l}},f)\,I(\overline{\mathbf{l}},f)\,W_f(f)\,e^{-i2\pi \overline{\mathbf{u}}\cdot\overline{\mathbf{l}}}\,\dif^2 \overline{\mathbf{l}}.
\end{align}

We define the {\it delay spectrum} $V_\eta(\overline{\mathbf{u}},\eta)$ to be the inverse Fourier transform of $V_f(\overline{\mathbf{u}},f)$ along the frequency coordinate:
\begin{align}\label{eqn:delay-transform}
  V_\eta(\overline{\mathbf{u}},\eta) &= \int V_f(\overline{\mathbf{u}},f)\,W'_f(f)\,e^{i2\pi\eta f}\,\dif f,
\end{align}
where, $W'_f(f)$ is a frequency window weighting function which can be chosen to control the quality of the delay spectrum \citep{thy13,ved12}. $\eta$ represents the signal delay between the antenna pairs in an interferometer given by:
\begin{equation}\label{eqn:delay}
  \eta = \frac{\overline{\mathbf{x}}\cdot\overline{\mathbf{l}}}{c} = \frac{\overline{\mathbf{u}}\cdot\overline{\mathbf{l}}}{f}.
\end{equation}
Note that as the visibility is a fully complex quantity, its Fourier transform is fully Nyquist sampled over positive and negative delays. Negative and positive delay\footnote{The terms delay and lag are used interchangeably in this paper.} correspond to the two hemispheres of the sky as transected by the antenna pair's baseline vector. For example, a perfectly Eastward oriented baseline vector will obesrve an object in the Eastern sky at positive delay, and conversely for an object in the Western sky.

Since we are studying a redshifted H{\sc i} spectral line, $f$ is a measure of cosmological distance along the line of sight. Consequently, $\eta$ is a measure of line of sight spatial frequency mode, $k_\parallel$. Thus, $|V_\eta(\overline{\mathbf{u}},\eta)|^2$ is directly related to the spatial power spectrum of redshifted H{\sc i} distribution, $|V(\overline{\mathbf{k}}_\perp,k_\parallel)|^2$. However, foreground objects emit in radio frequencies and contaminate the signals from the redshifted H{\sc i}. Due to their spatial and spectral properties, $V_\eta(\overline{\mathbf{u}},\eta)$ also contains contamination from foreground radio objects \citep{thy13,tro12,mor12,bow09}. These contaminations have to be characterized precisely in order to reduce their impact on redshifted H{\sc i} power spectrum detection sensitivity.

@@

The MWA was completed in 2012, and after commissioning, began its EoR observing program in 2013. In its final configuration the MWA is a 128--tile interferometer array capable of observing a 30~MHz band anywhere in the range 80--300~MHz. Each tile is a phased array of 16 dipoles each in the shape of a bow--tie characterized by a grating response with a $\gtrsim$~20\arcdeg wide primary field of view and multiple secondary side lobes. The tile arrangement is a centrally condensed core of $\sim$~300~m --- there are many spacings in the range 5--50~m --- and a radial density that falls off as the inverse of the radius with the longest baselines at 3~km \citep{bea12}. Here we focus on antenna spacings $|\overline{\mathbf{x}}| \le 200$~m (spatial scales relevant to reionization) and a frequency range 170--200~MHz.

@@

The LSTs of these two snapshots are shown in Figure~\ref{fig:pointings} as vertical lines at -1.92~hours (i.e., 22.08~hours) and 0.09~hours which are hereafter denoted as {\it off--zenith} and {\it zenith} pointings respectively. %These pointings (phased array power--patterns) are centered at RA = 4\fdg 387, Dec = -29\fdg 86 and RA = -28\fdg 8, Dec = -26\fdg 701, whereas the visibilities themselves are phased to zenith corresponding to RA = -28\fdg 8, Dec = -26\fdg 701, and RA = 1\fdg 35, Dec = -26\fdg 701 respectively.

@@

\subsubsection{Initial Data processing}\label{sec:data-analysis}

The data are flagged\footnote{\url{http://sourceforge.net/p/aoflagger}} for interference \citep{off12,off10}, removing 3\% of the data and averaged in time  and frequency from the raw 0.5~s, 40~kHz to 2s, 80~kHz. These data are then calibrated to a simulation of the sky containing 2420 point--like objects selected from the MWA Commissioning Survey (MWACS; Hurley-Walker et al, in prep). MWACS was compiled using observations taken with subsets of the array during the final buildout and commissioning. It has a flux density limit of 25~mJy and a declination range of -12\arcdeg to -40\arcdeg evenly covering the field of view of the observations reported here. The objects used in calibration are selected to lie inside the 5\% contour of the primary lobe of the tile power--pattern. The calibration algorithm computes complex gain solutions per channel per antenna averaged to two minute intervals. The solutions are still fairly low signal--to--noise so we iteratively project along the antenna and frequency dimensions to capture the relatively independent passband and antenna--to--antenna variation. First, we average the channel gains over all antenna to obtain a high signal--to--noise measurement of the bandpass. After applying this single passband, we do a second round of calibration and fit second and first order polynomials for amplitude and phase respectively for each antenna. This flattens any residual variation in bandpass and removes small phase slopes due to variation in cable delay. Finally, we fit for an additional phase known to be caused by reflection in 150~m length cables. %possibly could add elaboration, if I knew exactly which version of the cal was done here. doesn't really matter

@@

We obtain the delay spectrum of these calibrated visibilities taking the delay transform of each baseline's spectrum (equation~\ref{eqn:delay-transform}) choosing $W'_f(f)$ to be a {\it Blackman--Harris} window function. The spectrum is multiplied by  the additional weight of the flagged channels which occur rarely for interference and at regular spacing every 1.28~MHz where the edges of coarse channels are known to have a slight aliasing effect. In delay--space, these weights translate into a convolution by a ringing harmonic point spread function (PSF). We deconvolve this PSF using a one dimensional CLEAN algorithm \citep{tay99} as described for the delay axis by \citet{par09,par12}. The CLEAN procedure iteratively finds and subtracts peak values convolved by the Fourier transform of the weights. We limit the selection of peaks to modes inside the horizon delay limits corresponding to smooth spectrum objects in the visible sky hemisphere.

@@

In subsequent plots, we plot the amplitude of the delay spectrum, $|V(|\overline{\mathbf{x}}|,\eta)|$, sorted by baseline length, $|\overline{\mathbf{x}}|$. Note that visibilities from different baselines have not been averaged together in this analysis.

We show results of delay spectrum from MWA data after deconvolution along delay axis in figure~\ref{fig:fhd_data}. The principle feature is the wedge--shaped structure, the boundaries of which, marked by solid white lines, increases approximately ({\bf why exactly is it only approximate?}) linearly with light travel time corresponding to the baseline length. Note that the wedge is not filled out uniformly. Each baseline has a different orientation and resulting response to different objects in the sky. It is this variation that we are investigating.

In the context of EoR studies in Fourier space, the H{\sc i} power spectrum from the EoR is expected to decrease rapidly with $|\overline{\mathbf{k}}|$. Hence, the brightest EoR signal will occur on the shortest baselines and smallest delays. Thus the region of interest for EoR studies is just beyond the horizon delay limits (solid white lines) and inside the first harmonic on either side ($|\eta| \lesssim 0.78\,\mu$s) due to the coarse channel structure. This forms a small region inside the so-called {\it EoR window}.

The {\it off--zenith} pointing has emission notably higher than in the {\it zenith} pointing inside the wedge boundaries. This is shown later to be due to response of the array to the bright Galactic center and Galactic plane in the westward sky. Consequently, the contamination into the {\it EoR window} is also higher. Emission in the {\it zenith} pointing is more centrally concentrated due to phasing of power--pattern to the zenith, whereas the {\it off--zenith} pointing has a power--pattern phased eastward of zenith and is also responsive to emission from angles far away from zenith.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/multi_baseline_fhd_delay_spectrum_snapshots.eps}
\caption{ Delay spectrum power spectra of two 112s  Murchison Widefield Array observations of the Epoch of Reionization field at RA=0, Dec=-30. At top an {\it off--zenith}  pointing when tiles have significant side lobe response in the direction of the galactic center. At bottom a {\it zenith} pointing where side lobes are minimal and the galaxy has completely set. To get the the power spectrum each baseline is Fourier transformed into delay space ($y$-axis) and the amplitude plotted as a logarithmic color in order of baseline length ($x$-axis). The logarithmic color scale (shown at the right) is common to both panels. White lines mark the boundaries of the {\it foreground wedge} determined by the light travel time across each baseline. Smooth spectrum objects are physically limited to the region below this line. The various features within the wedge depend on the \emph{orientation} of the baseline, the side lobe structure of the power--pattern and the sky at that instant. \label{fig:fhd_data}}
\end{figure}

Despite delay deconvolution, we still see leakage beyond the cleaned regions indicating the inability of the deconvolution algorithm to fit the measurements perfectly. This is predominant at short antenna spacings where the number of delay modes within the horizon is lesser. This is because the maximum delay envelope (boundary of the wedge) consists of mixed emission from large portions of the sky into a narrow range of delays and the deconvolution algorithm does not have sufficient support in delay space to fit the delay spectrum. However, we wish to emphasize that the purpose of deconvolution is to rid the delay spectra of instrumental artifacts to the best extent possible in order to see the underlying signatures of foreground emission. It is not our intention in this paper to use the deconvolution as a foreground removal tool. Hence, our discussion and conclusions are valid despite imperfections in deconvolution.

@@

We describe power--pattern and foreground models we have used in modeling the instrument. %todo

\subsection{power--pattern}\label{sec:power_pattern}

The precise power--pattern of the MWA tile ($A(\overline{\mathbf{l}},f)$ in equation~\ref{eqn:obsvis}) is a subject of active investigation. The most precise model currently available is that of Sutinjo et al (2014; submitted?) which models the 4$\times$4 array of bow-tie dipoles as a mutually-coupled array with the overall power--pattern of the dipole modeled via finite element electromagnetic simulation. To speed simulation we find that a phased array of isotropic radiators at a height of 0.3~m above an infinite ground plane provides a very good approximation to the FEKO (or just aforementioned?) simulation. We also assume that each individual dipole signal has random delay fluctuations of rms 0.05~ns, a number in line with the known stability level of the analog signal chain.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/delta_array_powerpattern_0.3m_ground_custom.eps}
\caption{Models of MWA tile power--pattern in Right Ascension ($\alpha$) and Declination ($\delta$) coordinates for the off--zenith (top) and zenith pointings (bottom) at 185~MHz. An MWA tile is modeled as a 4$\times$4 array of isotropic radiators placed 0.3~m above the ground plane. Random fluctutations of rms 0.05~ns have been added to delay corrections during the phased addition of voltages from these isotropic radiators. The logarithmic color scale (shown at the right) is common to both panels. \label{fig:power_pattern}}
\end{figure}

\subsection{Foreground Model}\label{sec:foreground}
%todo
The MWA has a very wide field of view ($\gtrsim 20$\arcdeg at 185~MHz) and has non-zero response all the way to the horizon. As we have seen above, foreground objects contaminate the region inside the ``wedge'' with an amplitude that depends on the amplitude of power--pattern in that direction, with objects at the horizon being closest to the boundaries of the wedge.

The degree to which these objects leak power into spectral line of sight modes above the wedge, a region known as the ``EoR Window'', depends primarily on the brightness of foregrounds near the boundaries of the wedge \citep{thy13,pob13,ved12,par12}. Thus, it is important to consider an all--sky model for foreground objects in evaluating the features seen in the power spectrum instead of restricting to the primary field of view.

\citet{bea13} estimated the sensitivity of the MWA to EoR H{\sc i} power spectrum detection taking into account thermal noise effects in the {\it EoR window}. \citet{thy13} estimated the sensitivity taking into account the spillover from foreground contamination from unsubtracted extragalactic compact objects in the {\it EoR window} besides thermal noise. In this paper, we include the diffuse Galactic emission for enhancing our understanding of foreground signatures in the power spectrum. We use an all--sky foreground emission model that includes both diffuse and bright compact components.

@@

For the diffuse component, we use an all--sky radio foreground model \citep{deo08} to estimate the emission at 185~MHz. Since this map is predominantly based on the 408~MHz all--sky map of \citet{has82} which has an angular resolution of 0\fdg 85, we smoothed the 185~MHz diffuse emission map to the same resolution. However, to avoid any artifacts from sampling this map, we sample it at $\approx 27$\arcmin~intervals. Using the same model, we use the diffuse emission maps at 170~MHz and 200~MHz to estimate the spectral index map of diffuse emission.

A low resolution version of the diffuse foreground model used is shown in figure~\ref{fig:DSM}. Contours of the MWA tile power--pattern shown in figure~\ref{fig:power_pattern} are also overlaid. Of notable significance in the {\it off--zenith} pointing is the presence of a portion of the Galactic plane and the bright Galactic center in the visible hemisphere in the westward sky and the MWA tile's power receptivity is significantly high ($\gtrsim 12$\%) in that direction. In the {\it zenith} pointing, the Galactic plane has almost set and the contour level of power--pattern in that direction is at least 16 times smaller.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/dsm.eps}
\caption{Sky brightness temperature (in K) of the diffuse foreground model at 185~MHz in Right Ascension ($\alpha$) and Declination ($\delta$) coordinates visible during {\it off--zenith} (top) and {\it zenith} (bottom) pointings. The color scale on the right is logarithmic and is common to both panels. Contours of power--pattern shown in figure~\ref{fig:power_pattern} are overlaid. The contour levels shown are 0.00195, 0.00781, 0.0312, 0.125, and 0.5. The Galactic center and a portion of the Galactic plane are prominently visible during the {\it off--zenith} pointing and the MWA tile power receptivity is significant ($\gtrsim 12$\%) in that direction. In contrast, emission from the Galactic plane in {\it zenith} pointing is significantly lesser. \label{fig:DSM}}
\end{figure}
